\documentclass[12pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{array}

\title{\textbf{CSE 317: Design and Analysis of Algorithms} \\ Homework 2}
\author{Abdullah Khan Sherwani \\ ERP: 27880}
\date{}

\begin{document}
\maketitle

\section*{Question 1: Triangle Maximum Path Sum}

\subsection*{Problem Statement}
Consider a triangular array of integers with $n$ rows, where the $i$-th row contains exactly $i$ positive numbers. Starting from the top element, at each step from position $j$ in row $i$, we may move to position $j$ or $j+1$ in row $i+1$. The goal is to find a path from top to bottom that maximizes the sum of selected numbers.

\subsection*{Dynamic Programming Formulation}

\subsubsection*{Notation}
Let $T[i][j]$ denote the value at position $j$ in row $i$ of the triangle, where $1 \le i \le n$ and $1 \le j \le i$.

\subsubsection*{Subproblem Definition}
Define $\text{dp}[i][j]$ as the \textbf{maximum path sum} starting from the top of the triangle $(T[1][1])$ and ending at position $j$ in row $i$.

\subsubsection*{Base Case}
\[
    \text{dp}[1][1] = T[1][1]
\]
There is only one element in the first row, so the maximum sum to reach it is just its own value.

\subsubsection*{Recurrence Relation}
For row $i$ ($2 \le i \le n$) and position $j$ ($1 \le j \le i$):

\[
    \text{dp}[i][j] = T[i][j] + \begin{cases}
        \text{dp}[i-1][1] & \text{if } j = 1 \\[6pt]
        \text{dp}[i-1][j-1] & \text{if } j = i \\[6pt]
        \max\!\big(\text{dp}[i-1][j-1],\; \text{dp}[i-1][j]\big) & \text{if } 1 < j < i
    \end{cases}
\]

\textbf{Justification:} To arrive at position $j$ in row $i$, we must have come from position $j-1$ or position $j$ in row $i-1$ (since a parent at position $k$ can move to $k$ or $k+1$). Therefore:
\begin{itemize}[nosep]
    \item If $j = 1$: the only parent is position $1$ in row $i-1$.
    \item If $j = i$: the only parent is position $j-1 = i-1$ in row $i-1$.
    \item Otherwise: we take the maximum of the two possible parents.
\end{itemize}

\subsubsection*{Optimal Value}
The answer is:
\[
    \text{Maximum Path Sum} = \max_{1 \le j \le n} \text{dp}[n][j]
\]
We scan the last row of the DP table and return the largest value.

\subsection*{Algorithm}

\begin{algorithm}[H]
\caption{\textsc{TriangleMaxPathSum}$(T, n)$}
\begin{algorithmic}[1]
\Require Triangle $T[1..n][1..i]$ and number of rows $n$
\Ensure Maximum path sum from top to bottom
\State $\text{dp}[1][1] \gets T[1][1]$
\For{$i \gets 2$ \textbf{to} $n$}
    \For{$j \gets 1$ \textbf{to} $i$}
        \If{$j = 1$}
            \State $\text{dp}[i][j] \gets T[i][j] + \text{dp}[i-1][1]$
        \ElsIf{$j = i$}
            \State $\text{dp}[i][j] \gets T[i][j] + \text{dp}[i-1][j-1]$
        \Else
            \State $\text{dp}[i][j] \gets T[i][j] + \max(\text{dp}[i-1][j-1],\; \text{dp}[i-1][j])$
        \EndIf
    \EndFor
\EndFor
\State \Return $\max_{1 \le j \le n} \text{dp}[n][j]$
\end{algorithmic}
\end{algorithm}

\subsection*{Complexity Analysis}

The algorithm has \textbf{time complexity} $O(n^2)$ since it fills an $n \times n$ triangular DP table with constant-time operations per cell. The \textbf{space complexity} is $O(n^2)$ for storing the DP table, though it can be optimized to $O(n)$ by using only two rows at a time since each row depends only on the previous row.

\newpage

\section*{Question 2: Coin Change Problem}

\subsection*{Problem Statement}
Given a set of coin denominations $C = \{c_0, c_1, \ldots, c_{n-1}\}$ where each $c_i$ is a positive integer, and assuming each coin denomination can be used an unlimited number of times, determine the minimum number of coins required to make a target value $V$ exactly. If no combination of coins can form the target value $V$, report that no solution exists.

\subsection*{Dynamic Programming Formulation}

\subsubsection*{Notation}
Let $C[i]$ denote the value of the $i$-th coin denomination (0-indexed), where $0 \le i < n$.

\subsubsection*{Subproblem Definition}
Define $\text{dp}[v]$ as the \textbf{minimum number of coins} required to make the value $v$, for $0 \le v \le V$.

\subsubsection*{Base Case}
\[
    \text{dp}[0] = 0
\]
No coins are needed to make a value of 0.

\subsubsection*{Recurrence Relation}
For each value $v$ where $1 \le v \le V$:

\[
    \text{dp}[v] = \begin{cases}
        \infty & \text{if no coin denomination can form } v \\[6pt]
        1 + \min_{i: C[i] \le v} \text{dp}[v - C[i]] & \text{otherwise}
    \end{cases}
\]

\textbf{Justification:} To form value $v$, we can use any coin $C[i]$ such that $C[i] \le v$, and then recursively form the remaining value $v - C[i]$. We choose the coin that results in the minimum total number of coins.

\subsubsection*{Optimal Value}
The answer is:
\[
    \text{Minimum Coins} = \begin{cases}
        \text{dp}[V] & \text{if } \text{dp}[V] < \infty \\
        \text{No solution exists} & \text{otherwise}
    \end{cases}
\]

\subsection*{Algorithm}

\begin{verbatim}
def min_coins(C, V):
    """
    Find the minimum number of coins required to make value V.
    """
    # dp[v] = minimum coins needed to make value v
    dp = [float('inf')] * (V + 1)
    dp[0] = 0  # Base case: 0 coins for value 0

    # Fill DP table for each value from 1 to V
    for v in range(1, V + 1):
        # Try using each coin denomination
        for coin in C:
            if coin <= v and dp[v - coin] != float('inf'):
                dp[v] = min(dp[v], 1 + dp[v - coin])

    # Return result
    return dp[V] if dp[V] != float('inf') else -1
\end{verbatim}

\subsection*{Complexity Analysis}

The algorithm has \textbf{time complexity} $O(V \cdot n)$ where $V$ is the target value and $n$ is the number of coin denominations. The outer loop iterates $V$ times, and for each value, we examine all $n$ coin denominations. The \textbf{space complexity} is $O(V)$ for storing the DP table that tracks the minimum coins needed for each value from 0 to $V$.

\end{document}
